---
title: "Microbial Community Analysis workshop"
output: 
  md_document:
    variant: markdown_github
---

# Using the Phyloseq package

## installation for bioconductor

```{r,eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("phyloseq")
library(phyloseq)
library(ggplot2)

```

## Read in the dataset, biom file

```{r}
slashpile_16sV1V3 <- "16sV1V3.biom"
s16sV1V3 = import_biom(BIOMfilename = slashpile_16sV1V3, parseFunction = parse_taxonomy_default)
colnames(tax_table(s16sV1V3)) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")
rank_names(s16sV1V3)

s16sV1V3
```

## Filtering

```{r}
prevelancedf = apply(X = otu_table(s16sV1V3),
                 MARGIN = 1,
                 FUN = function(x){sum(x > 0)})
# Add taxonomy and total read counts to this data.frame
prevelancedf = data.frame(Prevalence = prevelancedf,
                      TotalAbundance = taxa_sums(s16sV1V3),
                      tax_table(s16sV1V3))
prevelancedf
```

The following ensures that features with ambiguous phylum annotation are also removed. Note the flexibility in defining strings that should be considered ambiguous annotation.

```{r}
s16sV1V3.1 <- subset_taxa(s16sV1V3, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized"))
s16sV1V3.1
```

### Now subset low occurance taxa

```{r}
plyr::ddply(prevelancedf, "Phylum", function(df1){
  data.frame(mean_prevalence=mean(df1$Prevalence),mean_abundance=mean(df1$TotalAbundance,na.rm = T),stringsAsFactors = F)
  })
```
### Define phyla to filter
```{r}
phyla2Filter = c("p__Aquificae", "p__candidate division ZB3",
  "p__Crenarchaeota","p__Deinococcus-Thermus","p__Omnitrophica","p__Tenericutes","p__Thermodesulfobacteria")
# Filter entries with unidentified Phylum.
s16sV1V3.2 = subset_taxa(s16sV1V3.1, !Phylum %in% phyla2Filter)
s16sV1V3.2
```

### Subset to the remaining phyla
```{r}
prevelancedf1 = subset(prevelancedf, Phylum %in% get_taxa_unique(s16sV1V3.2, taxonomic.rank = "Phylum"))
ggplot(prevelancedf1, aes(TotalAbundance, Prevalence / nsamples(s16sV1V3.2),color=Phylum)) +
  # Include a guess for parameter
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

```{r}
#  Define prevalence threshold as 5% of total samples
prevalenceThreshold = 0.10 * nsamples(s16sV1V3.2)
prevalenceThreshold

## [1] 18

# Execute prevalence filter, using `prune_taxa()` function
keepTaxa = rownames(prevelancedf1)[(prevelancedf1$Prevalence >= prevalenceThreshold)]
s16sV1V3.3 = prune_taxa(keepTaxa, s16sV1V3.2)
s16sV1V3.3
```